Musical elements in the discrete-time representation of sound

### 2017-07-29 article/acm/body.tex
revisada parte de harmonia

For general guidance:
\begin{itemize}
        \item A fifth confirms the root (fundamental).
		There are theoretical discussion about why this happens, and the most usual arguments are that the fifth is the first (non-octave) harmonic of a note and that the harmonics of the fifth are in the harmonics of the fundamental.
		Important here is to grasp the fact that musical theory and practice assures that the fifth establishes the fundamental as the root of a chord.
        \item Major or minor thirds from the root entails major or minor chord qualities.
        \item Every tritone, especially if built between a major third and a minor seventh, tends to resolve into a third or a sixth.
        \item Note duplication is avoided. If duplication is needed, the preference is, in descending order: the root, fifth, third and seventh.
        \item Note omission is avoided in the triad. If needed, the fifth is first considered for omission, then third and then the fundamental.
        \item It is possible to build chords with notes different from triads, particularly if they obey a recurrent logic or musical concatenation that justifies these different notes.
        \item Chords built by successive intervals different from thirds – like fourths and seconds -- are recurrent in compositions of advanced tonalism or experimental character.
        \item The repetition of chord successions (or their characteristics) fixes a trajectory and makes it possible to 
introduce exotic formations without implying in musical incoherence.
\end{itemize}

\subsection{Atonal and tonal harmonies, harmonic expansion and modulation}\label{subsec:harmonia}

Omission of basic tonal chaining is the key to obtain modal and atonal harmonies. In the absence of these minimal tonal structures, 
harmony is considered modal if the notes match with some diatonic scale (see equations~\ref{eq:escalas}) or if notes are presented in a small number. If basic tonal chainings are absent and notes do not match any diatonic scale and are diverse and dissonant (by relation with each other) enough to avoid reduction by polarization, harmony is atonal. In this classification, the modal harmony is not tonal or atonal and is reduced to the incidence of notes within the diatonic scale (or simplifications) and to the absence of tonal structures. Following this concept, one observes that atonal harmony is hard to be realized and, indeed, no matter how dissonant and diverse
a set of notes is, tonal harmonies arise if not avoided~\cite{harmEXT}.

\subsubsection{Atonal harmony}

In fact, the techniques around atonal music aim at structures for avoiding direct relation with modes and tonality. Manifesting such structures is of such difficulty that the dodecafonism arouse. The purpose of dodecafonism is to use a set of notes (ideally 12 notes), and to perform each note, one by one, in the same
order. In this context, the tonic becomes difficult to be established. Nevertheless, the western listener searches for tonal traces in music and easily finds them by unexpected and tortuous paths. The use of dissonant intervals (especially tritones) without resolution, reinforces the absence of tonality. In this context, while creating a music
piece, it is possible to:

\begin{itemize}
     \item Repeat notes. By considering immediate repetition as an extension of previous incidence, the use of the same note in sequence does not add relevant information.
     \item To play adjacent notes at the same time, making harmonic intervals and chords.
     \item Present durations and pauses with liberty, respecting notes order.
     \item Vary by extension, transposition, translation, inversion, retrograde and retrograde inversion. See subsections~\ref{subsec:motivos}
     and~\ref{subsec:usosmusicais3} for more details.
     \item Accounting for the presence of note structures, it is possible to use variations in orchestration, articulation, spatialization, among others.
\end{itemize}

The atonal harmony can be observed, paradigmatically, within these presented conditions (a simple dodecaphonic model). Most of what was written by great dodecafonic composers, 
e.g. Alban Berg and even Schoenberg, had the purpose of mixing tonal and atonal techniques.

\subsubsection{Tonal harmony}

In the XX century, rhythmic music and with emphasis on sonorities/timbres, extended the concepts of tonality and harmony. Hence, tonal harmony has strong influence on artistic movements and commercial venues. In addition, dodecafonism itself is considered of tonal nature because it denies tonal characteristics of polarization.
In tonal or modal music, chords -- like the ones listed in
equations~\ref{triades}-- built on the root note of each degree of a scale  -- such as listed in equations~\ref{eq:escalas} --  form the pilars of the harmonic field. Music harmony aims at studying the incidence of chord progressions and chaining rules. 
Even a monophonic melody generates harmonic fields, making it possible to observe the suggested chords at each passage.

In the traditional tonal music, a scale has its tonic (first
degree) on any note, and can be major (with the same notes of Ionian mode) or minor (same notes as Eolian mode, 'natural minor', which has both harmonic and melodic versions, see equations~\ref{eq:escalasMenores}). The
scale is the base for triads, each with its root in a degree: $\hat{1},\hat{2},\hat{3},\hat{4},\hat{5},\hat{6},\hat{7}$. 
To build triads, the third and the fifth notes above the root are taken together with the root (or fundamental).
$\hat{1},\hat{3},\hat{5}$ is the first degree chord,
 built on top of scale's first degree and central for tonal music. The chords of the fifth degree $\hat{5},\hat{7},\hat{2}$ ($\hat{7}$ sharp when a minor scale) and of the forth degree $\hat{4},\hat{6},\hat{1}$ are secondary. After these, other degrees are considered. The `traditional harmony' comprises conventions and stylistic techniques to create chainings with such chords~\cite{Harmonia}.

The `functional harmony' ascribes functions to these three central chords and tries to understand their use by means of these functions. The chord built on top of the first degree is the \textbf{tonic} chord (\textit{T} or \textit{t} for a major or minor tonic, respectively) and its function (role) consists of maintaining a center, usually referred to as a ``ground'' for the music. The chord built on the fifth degree is the \textbf{dominant} (\textit{D}, the dominant is always major) and its function is to lean for the tonic
(the dominant chord asks for a conclusion and this conclusion is the tonic). Thus, the dominant chord guides the music to the tonic. The triad built under the forth degree is the \textbf{subdominant} (\textit{S} or \textit{s} for a major or minor
subdominant, respectively) and its function is to deviate the music from the tonic. The process aims at confirming the tonic using tonic-dominant-tonic chains which are expanded by using other chords in different ways.

The remaining triads are associated to these three most important chords. In the major scale, the associated relative (relative tonic \textit{Tr}, relative
subdominant \textit{Sr} and relative dominant \textit{Dr}) is the triad built a third below, and the associated counter-relative (counter-relative tonic \textit{Ta}, counter-relative subdominant \textit{Sa} and the counter-relative dominant \textit{Da}) is the triad built in a third above. In the minor scale the same happens, but the triad a third below is called counter-relative (tA, sA) and the triad a third above is called relative (tR,
sR). The precise functions and musical effects of these chords are
controversial. Table~\ref{tab:harmonia} shows relations between the triads built in each degree of the major scale.

\begin{table}[htp!]
\centering
\caption{Summary of tonal harmonic functions of the major scale. Tonic is the music center, the dominant goes to the tonic and the subdominant moves the music away from
the tonic. The three chords can, at first, be freely replaced by their
respective relative and counter-relatives.}
\begin{tabular}{l | c | r}
relative & main chord of the function & counter-relative \\\hline\hline
$\hat{6},\hat{1},\hat{3}$ & tonic:       $\hat{1},\hat{3},\hat{5}$ & $\hat{3}, \hat{5},      \hat{7}$ \\
$\hat{3},\hat{5},\hat{7}$ & dominant:    $\hat{5},\hat{7},\hat{2}$ & [ $\hat{7},\hat{2},\hat{4}\#$ ] \\
$\hat{2},\hat{4},\hat{6}$ & subdominant: $\hat{4},\hat{6},\hat{1}$ & $\hat{6},\hat{1},       \hat{3}$
\end{tabular}
\label{tab:harmonia}
\end{table}

The dominant counter-relative should form a minor chord. It explains the change in the forth degree by a semitone above $\hat{7}\#$. The diminished chord
$\hat{7},\hat{2},\hat{4}$, is generally considered a `dominant seventh chord with omitted root'~\cite{Koellheuteur}.
In the minor mode, there is a change in $\hat{7}$ by an ascending semitone to yield a unique semitone separating $\hat{7}$ and $\hat{1}$, and making the dominant possible (that should be major and goes to the tonic). In this way, the dominant is always major, for both major and minor scale and, therefore, even in a minor tone the relative dominant remains a third below, and in the counter-relative, a third above.

\subsubsection{Tonal expansion: individual functions and chromatic mediants}

%% dominante e subdominante individualmente, ou, dominante e subdominante individual?

Each one of these chords can be confirmed and developed by performing their individual dominant or subdominant, which are the triads based
on a fifth above or a fifth below, respectively. These individual dominants and subdominants,
in the same way, have also subdominants and individual dominants of their own. Given a tonality, any chord can occur, no matter
how distant it is from the harmonic field and from the notes within the scale. The unique condition is that the occurrence presents a coherent trajectory of dominants and subdominants to the original tonality.

Mediants, or 'chromatic mediants', are two for each chord: the upper
mediant, formed by the root at the third of original chord; and the lower mediant, formed by the fifth at the third of original chord.
Both chords also are formed by a third, but with a chromatic alteration regarding the original chord. If two chromatic alterations exist, i.e.\ two notes altered by one semitone each regarding the original chord, it is a 'doubly chromatic mediant'. Again, there are two forms for each chord: the upper form, with a third in the fifth of the original triad; and the lower form, with a third in the root
of the original triad. It is worth observing that a major chord has major chromatic and doubly chromatic mediants. A minor chord has minor chromatic and doubly chromatic mediants. (Recall that relatives and counter-relatives have opposite major-minor quality.)
This relation between chords is considered advanced tonalism, sometimes even considered expansion and dissolution of tonalism,
with strong and impressing effects although they are perfectly consonant triads. Chromatic mediants are used since the end of Romanticism by Wagner, Lizt, Richard Strauss,
among others, and are quite simple to be realized~\cite{Harmonia,Salzer}.

\subsubsection{Modulation}

Modulation is the change of key (tonic, or tonal center) in a music, being characterized by start and end keys, and transition artifacts.
Keys are always conceived as related by fifths and their relative and counter-relatives. Some ways to perform modulation include:

\begin{itemize}
    \item Transposing the discourse to a new key, without any preparation. It is a common Baroque procedure although incident at other periods as well. Sometimes it is 
    called phrasal modulation or unprepared modulation.
    \item  Careful use of an individual dominant, and perhaps also the individual
    subdominant, to confirm change in key and harmonic field.
    \item Use of chromatic alterations to reach a chord in the new key starting from a chord in the previous key. Called chromatic modulation.
    \item Featuring a unique note, possibly repeated or suspended with no accompaniment, common to start and end keys, it constitutes a peculiar way
    to introduce the new harmonic field.
    \item Changing the function, without changing the notes, of a chord to
    contemplate a new key. This procedure is called enharmony.
    \item Maintaining the tonal center and changing the key quality from major to minor
    (or vice-verse) comprehends a parallel modulation. Keys with same tonic and
    different quality are known as homonyms.
\end{itemize}

The dominant has great importance and is a natural pivot into modulations,
which leads to the circle of fifths~\cite{Harmonia,Salzer,Koellheuteur,Harmony}.
Other inventive ways to modulate are possible, to point but one common example, the minor thirds tetrad ($E_i^{tm}$ in equations~\ref{escSim}) can be sustained to bridge to other tonalities, with the ease of its both tritones.
The music piece \emph{Acorde cedo} explores these chord relations, and is implmemented online as part of \massa\ toolbox~\cite{MASSA}.

>>>>

For general guidance:
\begin{itemize}
        \item A fifth confirms the root (fundamental).
		There are theoretical discussion about why this happens, and the most usual arguments are that the fifth is the first (non-octave) harmonic of a note and that the harmonics of the fifth are in the harmonics of the fundamental.
		Important here is to grasp the fact that musical theory and practice assures that the fifth establishes the fundamental as the root of a chord.
        \item Major or minor thirds from the root entails major or minor chord qualities.
        \item Every tritone, especially if built between a major third and a minor seventh, tends to resolve into a third or a sixth.
        \item Note duplication is avoided. If duplication is needed, the preference is, in descending order: the root, fifth, third and seventh.
        \item Note omission is avoided in the triad. If needed, the fifth is first considered for omission, then third and then the fundamental.
        \item It is possible to build chords with notes different from triads, particularly if they obey a recurrent logic or musical concatenation that justifies these different notes.
        \item Chords built by successive intervals different from thirds – like fourths and seconds -- are recurrent in compositions of advanced tonalism or experimental character.
        \item The repetition of chord successions (or of characteristics they hold) fixes a trajectory and makes it possible to 
introduce exotic arrangements without implying musical incoherence.
\end{itemize}

\subsection{Atonal and tonal harmonies, harmonic expansion and modulation}\label{subsec:harmonia}
Omission of basic tonal structures is the key to achieving modal and atonal harmonies. In the absence of minimal tonal organization,
harmony is (usually) considered modal if the notes match with some diatonic scale (see equations~\ref{eq:escalas}) or if there is only a small number of notes. If basic tonal progressions are absent and notes do not match any diatonic scale and are enough diverse and dissonant (between themselves) to avoid reduction of the notes by polarization\footnote{By polarization we mean having some notes that are way more important than others and to which the other notes are ornaments or subordinates.}, the harmony is atonal. In this classification, the modal harmony is not tonal or atonal and is reduced to the incidence of notes within a (most often diatonic) scale and to the absence of tonal structures. Following this concept, one observes that atonal harmony is hard to be realized and, indeed, no matter how dissonant and diverse
a set of notes is, tonal harmonies arise very easily if not avoided~\cite{harmEXT}.

\subsubsection{Atonal harmony}
In fact, atonal music techniques avoid that a direct relation of the notes with modes and tonality be established. Manifesting such atonal structures is of such difficulty that the dodecafonism emerged. The purpose of dodecafonism is to use a set of notes (ideally 12 notes), and to perform each note, one by one, in the same
order. In this context, the tonic becomes difficult to be established. Nevertheless, the western listener automatically searches for tonal elements in music and obstinately finds them by unexpected and tortuous paths. The use of dissonant intervals (especially tritones) without resolution reinforces the absence of tonality. In this context, while creating a musical
piece, it is possible to:
\begin{itemize}
     \item Repeat notes. By considering immediate repetition as an extension of the previous incidence, the use of the same note in sequence does not add relevant information.
     \item To play adjacent notes (e.g. of a dodecafonic progression) at the same time, making harmonic intervals and chords.
     \item Use durations and pauses with liberty, respecting notes order.
     \item Vary note sequences by temporal magnification and translation or pitch transposition and by pitch sequence inversion, retrograde and retrograde inversion. See subsections~\ref{subsec:motivos}
     and~\ref{subsec:usosmusicais3} for what these terms mean.
     \item Make variations in orchestration, articulation, spatialization, among other possibilities in presenting the same notes.
\end{itemize}

The atonal harmony can be observed, paradigmatically, within these presented conditions (which is a simple dodecaphonic model). Most of what was written by great dodecafonic composers,
e.g. Alban Berg and even Schoenberg, had the purpose of mixing tonal and atonal techniques.
Most frequently, atonal music is not strictly dodecafonic, but "serial", i.e. they use the same kind of techniques based in a sequence of notes (called the series or row).

\subsubsection{Tonal harmony}
In the XX century, music with emphasis on sonorities/timbres and rhythm, extended the concepts of tonality and harmony. Even so, tonal harmony is very often in artistic movements and commercial venues. In addition, dodecafonism itself is sometimes considered of tonal nature because it was conceived to deny tonal characteristics of polarization.
In tonal or modal music, chords -- like the ones listed in
equations~\ref{triades}-- built on the root note of each degree of a scale  -- such as listed in equations~\ref{eq:escalas} --  form the pillars of harmony. Tonal (and modal) harmony aims at understanding the incidence of chord progressions and chaining rules.
Even a monophonic melody entails harmonic fields, making it possible to perceive the chord progression even in unaccompanied melodies.

In the traditional tonal music, a scale has its tonic (first
degree) on any note, and can be major (with the same notes of Ionian mode) or minor (same notes as Eolian mode, 'natural minor', which has both harmonic and melodic versions, as in equations~\ref{eq:escalasMenores}). The
scale is the base for triads, each with its root in a degree: $\hat{1},\hat{2},\hat{3},\hat{4},\hat{5},\hat{6},\hat{7}$.
To build triads, the third and the fifth notes above the root are taken together with the root (or fundamental).
$\hat{1},\hat{3},\hat{5}$ is the first degree chord,
 built on top of the scale's first degree and central for tonal music. The chords of the fifth degree $\hat{5},\hat{7},\hat{2}$ ($\hat{7}$ sharp when in a minor scale) and of the forth degree $\hat{4},\hat{6},\hat{1}$ are also important. The triads build on the other degrees less important then these and are usually understood in relation to them. The `traditional harmony' comprises conventions and stylistic techniques to create progressions with such chords~\cite{Harmonia}.

The `functional harmony' ascribes functions to the three main chords and tries to understand their use by means of these functions. The chord built on top of the first degree is the \textbf{tonic} chord (\textit{T} or \textit{t} for a major or minor tonic, respectively) and its function (role) consists on maintaining a center, usually referred to as a ``ground'' for the music. The chord built on the fifth degree is the \textbf{dominant} (\textit{D}, the dominant is always major) and its function is to lean for the tonic
(the dominant chord asks for a conclusion and this conclusion is the tonic). Thus, the dominant chord guides the music to the tonic. The triad built under the fourth degree is the \textbf{subdominant} (\textit{S} or \textit{s} for a major or minor
subdominant, respectively) and its function is to deviate the music from the tonic. The tonal discourse aims at confirming the tonic using tonic-dominant-tonic chains which are expanded by using other chords in numerous ways.

The remaining triads are associated to these three most important chords. In the major scale, the associated relative (relative tonic \textit{Tr}, relative
subdominant \textit{Sr} and relative dominant \textit{Dr}) is the triad built a third below, and the associated counter-relative (counter-relative tonic \textit{Ta}, counter-relative subdominant \textit{Sa} and the counter-relative dominant \textit{Da}) is the triad built in a third above. In the minor scale the same happens, but the triad a third below is called counter-relative (tA, sA) and the triad a third above is called relative (tR,
sR). The precise functions and musical effects of these chords are
controversial but are basically the same as the chords they are associated to. Table~\ref{tab:harmonia} shows relations between the triads built in each degree of the major scale.

\begin{table}[htp!]
\centering
\caption{Summary of tonal harmonic functions on the major scale. Tonic is the musical center, the dominant goes to the tonic and the subdominant moves the music away from
the tonic. The three chords can, in principle, be freely replaced by their
respective relative or counter-relatives.}
\begin{tabular}{l | c | r}
relative & main chord of the function & counter-relative \\\hline\hline
$\hat{6},\hat{1},\hat{3}$ & tonic:       $\hat{1},\hat{3},\hat{5}$ & $\hat{3}, \hat{5},      \hat{7}$ \\
$\hat{3},\hat{5},\hat{7}$ & dominant:    $\hat{5},\hat{7},\hat{2}$ & [ $\hat{7},\hat{2},\hat{4}\#$ ] \\
$\hat{2},\hat{4},\hat{6}$ & subdominant: $\hat{4},\hat{6},\hat{1}$ & $\hat{6},\hat{1},       \hat{3}$
\end{tabular}
\label{tab:harmonia}
\end{table}

The dominant counter-relative should form a minor chord. It explains the change in the forth degree by a semitone above $\hat{4}\#$. The diminished chord
$\hat{7},\hat{2},\hat{4}$, is generally considered a `dominant seventh chord with the root omitted'~\cite{Koellheuteur}.
In the minor mode, there is a change in $\hat{7}$ by an ascending semitone to achieve a separation between $\hat{7}$ and $\hat{1}$ of a semitone. This is important for the dominant function (which should lean to the tonic). In this way, the dominant is always major, for both major and minor scale and, therefore, even in a minor tone the relative dominant remains a third below, and the counter-relative remains a third above.

\subsubsection{Tonal expansion: individual functions and chromatic mediants}
Each chord can be stressed and developed by performing their individual dominant or subdominant, which are the triads based
on a fifth above or a fifth below, respectively. These individual dominants and subdominants,
in the same way, have also subdominants and individual dominants of their own. Given a tonality, any chord can occur, no matter
how distant it is from the most basic chords and from the notes of the scale. The unique condition is that the occurrence presents a coherent trajectory of dominants and subdominants (or their relatives and counter-relatives) to the original tonality.

There are two mediants, or 'chromatic mediants', for each chord: the upper
mediant, formed with the root at the third of original chord; and the lower mediant, formed by the fifth at the third of original chord.
Both chords also are formed by a third, but with a chromatic alteration regarding the original chord. If two chromatic alterations exist, i.e.\ two notes altered by one semitone each regarding the original chord, it is a 'doubly chromatic mediant'. Again, there are two forms for each chord: the upper form, with a third in the fifth of the original triad; and the lower form, with a third in the root
of the original triad. A major chord has major chromatic and doubly chromatic mediants. A minor chord has minor chromatic and doubly chromatic mediants. (Recall that relatives and counter-relatives have opposite major-minor quality.)
This relation between chords is considered advanced tonalism, sometimes even considered as an expansion and dissolution of tonalism,
with strong and impressive effects although they are simple, consonant triads. Chromatic mediants are used since the end of Romanticism by Wagner, Lizt, Richard Strauss, among others~\cite{Harmonia,Salzer}.

\subsubsection{Modulation}
Modulation is the change of key (tonic, or tonal center) in music, being characterized by start and end keys, and transition artifacts.
Keys are always (thought of as) related by fifths and their relatives and counter-relatives. Some ways to perform modulation include:

\begin{itemize}
    \item Transposing the discourse to a new key, without any preparation. It is a common Baroque procedure although incident in other periods as well. Sometimes it is 
    called phrasal modulation or unprepared modulation.
    \item Careful use of an individual dominant, and perhaps also the individual
    subdominant, to confirm change in key and harmonic field.
    \item Use of chromatic alterations to reach a chord in the new key by starting from a chord in the previous key. Called chromatic modulation.
    \item Featuring a unique note, possibly repeated or suspended with no accompaniment, common to start and end keys, it constitutes a peculiar way
    to introduce the new harmonic field.
    \item Changing the function, without changing the notes, of a chord to
    contemplate a new key. This procedure is called enharmony.
    \item Maintaining the tonal center and changing the key quality from major to minor
    (or vice-versa) is a `parallel modulation'. Keys with same tonic but
		different (major/minor) qualities are known as homonyms.
\end{itemize}

The dominant has great importance and is a natural pivot in modulations,
a fact that leads to the circle of fifths~\cite{Harmonia,Salzer,Koellheuteur,Harmony}.
Other inventive ways to modulate are possible, to point but one common example, the minor thirds tetrad ($E_i^{tm}$ in equations~\ref{escSim}) can be sustained to bridge to other tonalities, with the ease of its both tritones.
The music piece \emph{Acorde cedo} explores these chord relations, and is implemented online as part of \massa\ toolbox~\cite{MASSA}.

### 2017-07-29 article/acm/foobody.tex
feita revisao de escalas e acordes
revisada parte de harmonia

\subsubsection{Chords}\label{subsec:acordes}

The simultaneous occurrence of three or more notes is observed by means of chords. Chords are based on triads, especially in tonal music. Triads are built by two successive thirds
within 3 notes: root, third and fifth. If the lower note of a chord is different from the root, this chord is an inverted chord as opposed to the chord in its root position. A closed position is any in which no chord note fits between two consecutive
notes~\cite{Lacerda}, any non-closed position is an open position. In closed and fundamental position,
and with fundamental in $0$, triads are described as:

\begin{equation}\label{triades}
\begin{split}
\text{major triad} = A_i^M= \{a_i^M\}_0^2=\{0,4,7\} \\ 
\text{minor triad} = A_i^m = \{a_i^m\}_0^2=\{0,3,7\} \\
\text{diminished triad} = A_i^d = \{a_i^d\}_0^2=\{0,3,6\} \\
\text{augmented triad} = A_i^a = \{a_i^a\}_0^2=\{0,4,8\}
\end{split}
\end{equation}

@@

For general guidance:

\begin{itemize}
        \item A fifth confirms the root (fundamental).
        \item Major or minor thirds suggest major or minor chord qualities.
        \item Every tritone, especially if built between a major third and a minor seventh, tends to resolve into a third or a sixth.
        \item Note duplication is avoided. If duplication is needed, the preferred order is: the root, fifth, third and seventh.
        \item Note omission is avoided in the triad. If needed, the fifth is first considered for omission, than third and, if needed, the fundamental.
        \item It is possible to build chords with notes different from triads, particularly if they obey a recurrent logic or musical concatenation that justifies these different notes.
        \item Chords built by successive intervals different from thirds – like fourths and seconds -- are recurrent in compositions of advanced tonalism or experimental character.
        \item The repetition of chord successions (or their characteristics) fixes a trajectory and makes it possible to 
introduce exotic formations without incoherence.
\end{itemize}

\subsection{Atonal and tonal harmonies, harmonic expansion and modulation}\label{subsec:harmonia}

Omission of basic tonal chaining is the key to obtain modal and atonal harmonies. In the absence of these minimal tonal structures, 
harmony is considered modal if the notes match with some diatonic scale (see equations~\ref{eq:escalas}) or if notes are presented in a small number. If basic tonal chainings are absent and notes do not match any diatonic scale and are diverse and dissonant (by relation with each other) enough to avoid reduction by polarization, harmony is atonal. In this classification, the modal harmony is not tonal or atonal and is reduced to the incidence of notes within the diatonic scale (or simplifications) and to the absence of tonal structures. Following this concept, one observes that atonal harmony is hard to be realized and, indeed, no matter how dissonant and diverse
a set of notes is, tonal harmonies arise if not avoided~\cite{harmEXT}.

\subsubsection{Atonal harmony}

In fact, the techniques around atonal music aim at structures for avoiding direct relation with modes and tonality. Manifesting such structures is of such difficulty that dodecafonism arises. The purpose of dodecafonism is to use a set of notes (ideally 12 notes), and to perform each note, one by one, in the same
order. In this context, the tonic becomes difficult to be established. Nevertheless, the western listener searches for tonal traces in music and easily finds them by unexpected and tortuous paths. The use of dissonant intervals (especially tritones) without resolution reinforces the absence of tonality. In this context, while creating a music
piece, it is possible to:

\begin{itemize}
     \item Repeat notes. By considering immediate repetition as an extension of previous incidence, the use of the same note in sequence does not add relevant information.
     \item To play adjacent notes at the same time, making harmonic intervals and chords.
     \item Present durations and pauses with liberty, respecting notes order.
     \item Vary by extension, transposition, translation, inversion, retrograde and retrograde inversion. See subsections~\ref{subsec:motivos}
     and~\ref{subsec:usosmusicais3} for more details.
     \item Accounting for the presence of note structures, it is possible to use variations in orchestration, articulation, spatialization, among others.
\end{itemize}

@@

\subsubsection{Tonal harmony}

In the XX century, rhythmic music and with emphasis on sonorities/timbres, extended the concepts of tonality and harmony. Hence, tonal harmony has strong influence on artistic movements and commercial venues. In addition, dodecafonism itself is considered of tonal nature because it denies tonal characteristics of polarization.
In tonal or modal music, chords -- like the ones listed in
equations~\ref{triades}-- built on the root note of each degree of a scale  -- such as listed in equations~\ref{eq:escalas} --  form the pilars of the harmonic field. Music harmony aims at studying the incidence of chord progressions and chaining rules. 
Even a monophonic melody generates harmonic fields, making it possible to observe the suggested chords at each passage.

@@

\subsubsection{Tonal expansion: individual functions and chromatic mediants}

%% dominante e subdominante individualmente, ou, dominante e subdominante individual?

Each one of these chords can be confirmed and developed by performing their individual dominant or subdominant, which are the triads based
on a fifth above or a fifth below, respectively. These individual dominants and subdominants,
in the same way, have also subdominants and individual dominants of their own. Given a tonality, any chord can occur, no matter
how distant it is from the harmonic field and from the notes within the scale. The unique condition is that the occurrence presents a coherent trajectory of dominants and subdominants to the original tonality.

@@

\subsubsection{Modulation}

Modulation is the change of key (tonic, or tonal center) in music, being characterized by start and end keys, and transition artifacts.
Keys are always conceived as related by fifths and their relative and counter-relatives. Some ways to perform modulation include:

>>>>

\subsubsection{Chords}\label{subsec:acordes}
The simultaneous occurrence of three or more notes is observed by means of chords. Chords are based on triads, especially in tonal music. Triads are built by two successive thirds
within 3 notes: root, third and fifth. If the lower note of a chord is different from the root, this chord is an inverted chord as opposed to the chord in its root position. A closed position is any in which no chord note fits between two consecutive
notes~\cite{Lacerda}, any non-closed position is an open position. In closed and fundamental position,
and with fundamental in $0$, triads are described as:
\begin{equation}\label{triades}
\begin{split}
\text{major triad} = A_i^M= \{a_i^M\}_0^2=\{0,4,7\} \\ 
\text{minor triad} = A_i^m = \{a_i^m\}_0^2=\{0,3,7\} \\
\text{diminished triad} = A_i^d = \{a_i^d\}_0^2=\{0,3,6\} \\
\text{augmented triad} = A_i^a = \{a_i^a\}_0^2=\{0,4,8\}
\end{split}
\end{equation}

@@

For general guidance:
\begin{itemize}
        \item A fifth confirms the root (fundamental).
        \item Major or minor thirds suggest major or minor chord qualities.
        \item Every tritone, especially if built between a major third and a minor seventh, tends to resolve into a third or a sixth.
        \item Note duplication is avoided. If duplication is needed, the preferred order is: the root, fifth, third and seventh.
        \item Note omission is avoided in the triad. If needed, the fifth is first considered for omission, than third and, if needed, the fundamental.
        \item It is possible to build chords with notes different from triads, particularly if they obey a recurrent logic or musical concatenation that justifies these different notes.
        \item Chords built by successive intervals different from thirds – like fourths and seconds -- are recurrent in compositions of advanced tonalism or experimental character.
        \item The repetition of chord successions (or their characteristics) fixes a trajectory and makes it possible to 
introduce exotic formations without incoherence.
\end{itemize}

\subsection{Atonal and tonal harmonies, harmonic expansion and modulation}\label{subsec:harmonia}
Omission of basic tonal chaining is the key to obtain modal and atonal harmonies. In the absence of these minimal tonal structures, 
harmony is considered modal if the notes match with some diatonic scale (see equations~\ref{eq:escalas}) or if notes are presented in a small number. If basic tonal chainings are absent and notes do not match any diatonic scale and are diverse and dissonant (by relation with each other) enough to avoid reduction by polarization, harmony is atonal. In this classification, the modal harmony is not tonal or atonal and is reduced to the incidence of notes within the diatonic scale (or simplifications) and to the absence of tonal structures. Following this concept, one observes that atonal harmony is hard to be realized and, indeed, no matter how dissonant and diverse
a set of notes is, tonal harmonies arise if not avoided~\cite{harmEXT}.

\subsubsection{Atonal harmony}
In fact, the techniques around atonal music aim at structures for avoiding direct relation with modes and tonality. Manifesting such structures is of such difficulty that dodecafonism arises. The purpose of dodecafonism is to use a set of notes (ideally 12 notes), and to perform each note, one by one, in the same
order. In this context, the tonic becomes difficult to be established. Nevertheless, the western listener searches for tonal traces in music and easily finds them by unexpected and tortuous paths. The use of dissonant intervals (especially tritones) without resolution reinforces the absence of tonality. In this context, while creating a music
piece, it is possible to:
\begin{itemize}
     \item Repeat notes. By considering immediate repetition as an extension of previous incidence, the use of the same note in sequence does not add relevant information.
     \item To play adjacent notes at the same time, making harmonic intervals and chords.
     \item Present durations and pauses with liberty, respecting notes order.
     \item Vary by extension, transposition, translation, inversion, retrograde and retrograde inversion. See subsections~\ref{subsec:motivos}
     and~\ref{subsec:usosmusicais3} for more details.
     \item Accounting for the presence of note structures, it is possible to use variations in orchestration, articulation, spatialization, among others.
\end{itemize}

@@

\subsubsection{Tonal harmony}
In the XX century, rhythmic music and with emphasis on sonorities/timbres, extended the concepts of tonality and harmony. Hence, tonal harmony has strong influence on artistic movements and commercial venues. In addition, dodecafonism itself is considered of tonal nature because it denies tonal characteristics of polarization.
In tonal or modal music, chords -- like the ones listed in
equations~\ref{triades}-- built on the root note of each degree of a scale  -- such as listed in equations~\ref{eq:escalas} --  form the pilars of the harmonic field. Music harmony aims at studying the incidence of chord progressions and chaining rules. 
Even a monophonic melody generates harmonic fields, making it possible to observe the suggested chords at each passage.

@@

\subsubsection{Tonal expansion: individual functions and chromatic mediants}
Each one of these chords can be confirmed and developed by performing their individual dominant or subdominant, which are the triads based
on a fifth above or a fifth below, respectively. These individual dominants and subdominants,
in the same way, have also subdominants and individual dominants of their own. Given a tonality, any chord can occur, no matter
how distant it is from the harmonic field and from the notes within the scale. The unique condition is that the occurrence presents a coherent trajectory of dominants and subdominants to the original tonality.

@@

\subsubsection{Modulation}
Modulation is the change of key (tonic, or tonal center) in music, being characterized by start and end keys, and transition artifacts.
Keys are always conceived as related by fifths and their relative and counter-relatives. Some ways to perform modulation include:
